Womb for One More? Fertility Expansion and the Limits of Early Health Investment: Empirical Evidence on the Quantity-Quality Trade-off from a Contraceptive Ban in Urban Philippines

### 2025-07-27 docs/paper.tex
revise theory and empirical sections in paper draft
Expanded and clarified the literature review, conceptual framework, and empirical strategy in the paper. Standardized terminology from 'quantity–quality' to 'quantity-quality', added new references and recent evidence, and restructured sections for improved flow. Updated title and subtitle for consistency across Rmd, tex, and PDF outputs.

\title{Womb for One More? Fertility Expansion and the Limits of Early
Health Investment}
\shortTitle{The Q-Q Trade-off in the Philippine Context}
% \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
% Surname2: affiliation2, address2, email2. Acknowledgements}}

@@

The quantity--quality (Q--Q) theory, a central idea in modern family
economics, holds that parents face a trade-off between the number of
children and the ``quality'' of investment---such as education or
health---they can provide to each. Quality in this context refers to the
human capital of each child: attributes like education, health, and
nutrition that enhance a child's future productivity and well-being. The
genesis of this idea traces back to Gary Becker's seminal work around
1960, which for the first time treated children as economic goods
subject to parental choice and budget constraints
\citep{becker1960economic}. Becker argued that as families become
wealthier, they may not simply want more children, but rather
better-raised children, much as a household might prefer a
higher-quality car or house over a greater quantity of them. This
proposition led to a formal theory in which increases in income or
changes in economic conditions cause parents to substitute child quality
for quantity, consistent with historical patterns of lower fertility and
higher educational attainment during economic development
\citep{galor2000population}.

@@

Becker and Lewis (1973) formalize the mutually reinforcing nature of the
quantity--quality cost structure. An increase in \(n\) raises the total
cost required to sustain a given level of \(q\) for each child, while a
higher level of \(q\) raises the marginal cost associated with having an
additional child. For example, allocating more resources to education or
health per child increases the financial burden of expanding family
size. This interdependence links the two decisions directly. The
household cannot choose \(n\) and \(q\) in isolation; each choice alters
the marginal cost of the other.

@@

In many dynastic models, child quality appears indirectly, often through
the child's future human capital or income. One variant assumes parents
value the aggregate human‐capital stock of their children. This
specification, combined with altruism, produces a similar trade‐off:
concentrating resources in fewer children raises each child's human
capital, which raises the dynasty's long‐run welfare. These
intergenerational extensions link micro‐level fertility decisions to
macroeconomic outcomes. By the late 1980s, work by Becker, Barro, and
others had recast fertility as an endogenous choice that interacts with
capital accumulation, income distribution, and policy. This laid the
foundation for unified growth theories, which view the quantity--quality
mechanism as central to demographic transition and long-run development.

@@

The unified growth literature places the Q--Q model within a more
general account of economic and demographic change. In this framework,
higher income or stronger returns to child quality reduce fertility and
help shift economies from stagnation toward sustained growth. Several
mechanisms support this transition, such as a fall in child labor, a
drop in child mortality, and a shift in parental priorities. These
models explain not only the presence of a quantity--quality trade-off
but also its rising influence at a specific point in history. The
evidence supports these claims: countries that saw earlier increases in
returns to education experienced earlier fertility decline, while delays
in reforms, such as public education or health access, corresponded to
prolonged high fertility. As a result, the Q--Q mechanism forms a key
component of unified growth theory.

@@

Contemporary research has further refined the quantity--quality model by
relaxing some of its initial simplifying assumptions. Three important
extensions involve (1) capital market imperfections that constrain
parents' ability to invest in child quality, (2) intra-household
conflict and bargaining between mothers and fathers over fertility
choices, and (3) recognition that child quality is multi-dimensional,
which extends beyond schooling to include health, nutrition, and other
facets of human capital.

To begin, I examine how credit constraints can give rise to poverty
traps. The canonical quantity--quality (Q--Q) framework assumes that
parents can reallocate resources freely; borrowing against future
earnings to finance schooling or health investments whenever the
expected return is high. In practice, especially in low-income settings,
credit markets function imperfectly: poor households typically cannot
secure loans to cover children's education or medical costs even when
such investments would yield substantial future gains. This market
failure magnifies the Q--Q trade-off. Becker, Lewis, and Willis--already
noted by \citet{grawe2008quality} as emphasizing ``resource
limitations''--implicitly recognized this issue, but contemporary models
make it explicit by imposing a borrowing constraint. Parents must fund
childrearing from current income alone; they cannot collateralize a
child's future wages to pay today's school fees. Consequently, when
income is low, each additional birth directly reduces the attainable
quality per child, potentially trapping families in a
low-income--high-fertility equilibrium.

Formal models show that ``the quality--quantity trade-off arises from a
binding credit constraint that prevents parents from borrowing against
future child income.'' Empirical work supports this mechanism.
\citet{kremer2002income} and \citet{de2003inequality} document that
countries facing tighter liquidity constraints tend to display higher
fertility and lower educational attainment, consistent with
liquidity-constrained parents favoring quantity over quality.
Cross-country evidence likewise indicates that where financial frictions
are more severe, the negative correlation between fertility and
schooling is stronger. Theoretically, introducing a borrowing limit can
generate multiple steady states: one with low fertility and high
investment when incomes suffice to cover quality costs, and another with
high fertility and minimal investment when they do not. Policy
instruments such as education subsidies or conditional cash transfers
effectively relax these constraints, which nudge households toward the
low-fertility, high-investment equilibrium. In short, incorporating
credit market imperfections deepens the explanatory power of the Q-Q
model: economic growth alone may not reduce fertility if households
remain too cash-poor to afford schooling, whereas targeted
quality-enhancing transfers can catalyze both demographic and
human-capital transitions.

A further refinement of the quantity--quality (Q--Q) model considers the
question of \emph{who} within the household makes fertility and child
investment decisions. The original Beckerian framework adopts a unitary
model of the family, and assumes a single utility function and complete
agreement between spouses over optimal fertility \(n\) and child quality
\(q\). In practice, however, empirical evidence reveals significant
heterogeneity in preferences between household members along gender
lines. According to \citet{oppenheim1987impact, thomas1990intra}, for
instance, men often desire more children than women and may differ in
their willingness to invest in each child's education or health. These
discrepancies have motivated game-theoretic models of intra-household
bargaining, in which fertility and investment outcomes reflect the
relative influence of each parent's preferences.

@@

The original Q--Q models often used a single catch-all variable for
child quality, typically thought of as education or ``expenditure per
child.'' Recent work emphasizes that child quality is multi-faceted, and
that parents make trade-offs along several dimensions of
investment---cognitive development, health, nutrition, etc. This is
particularly salient in developing countries, where basic health and
nutrition are pressing concerns alongside schooling.

@@

Beyond survival, nutritional status is a critical indicator of quality
in many low-income settings. Parents often face choices about how to
allocate food or medical care, especially under resource scarcity.
Empirical evidence supports the Q--Q model in this context: children
with many siblings often exhibit poorer nutritional outcomes. A recent
study using Vietnam Young Lives data finds that an additional sibling
reduces height-for-age and weight-for-age z-scores by about 0.3 standard
deviations on average \citep{chen2021population} This effect is
substantial, suggesting that children in larger families tend to be
shorter and lighter than their peers from smaller families, likely due
to the more limited distribution of healthcare and parental attention

Similar patterns emerge in sub-Saharan Africa
\citep{bishwakarma2019first} and South Asia , where first-born children
experience slower growth once younger siblings arrive---again pointing
to intrahousehold trade-offs. Theoretically, one can extend the Becker
model with a nutrition production function, where a child's health \(H\)
depends on food \(F\) and medical care \(M\), such that:

@@

Early empirical work in the Philippines provided suggestive support for
this theory. For example, an influential study by
\citet{horton1986child} used Philippine household data to jointly
examine fertility and child nutrition and treated nutritional status as
a measure of child quality. Horton found that better-educated mothers
and fathers tended to choose smaller families and achieved
better-nourished children--evidence of substitution away from
``quantity'' toward child ``quality.'' Notably, she observed strong
birth-order effects on nutrition (first-born children faring better than
later-born), which hint that parents may not evenly distribute quality
investments among all children. This early work indicated a
quantity--quality (Q--Q) trade-off in Philippine families, though it
largely documented correlations rather than definitive causation.

@@

The IV estimates confirmed a significant negative causal effect of
higher fertility on educational outcomes. \citet{orbeta2010number} found
that each additional child in the household reduced the proportion of
school-age children (6--24 years) attending school by roughly 19\% of
the baseline attendance rate. The trade-off was especially pronounced at
higher education levels: for example, the estimated drop in school
attendance was about 26\% at the secondary level and 57\% at the
tertiary level for each additional sibling. These are sizable effects,
implying that children from large families are substantially less likely
to remain in school, presumably due to tighter household budget
constraints or diluted parental attention. Moreover, the burden of the
trade-off appeared regressive: Orbeta's results showed much larger
schooling deficits from an extra child in poorer households than in
richer ones. For instance, in the poorest quintile, an additional
sibling reduced school attendance by an estimated 24\% (for ages 6--24),
compared to a 16\% reduction in the richest quintile. This regressive
pattern aligns with Becker's theory that resource constraints bind more
tightly for low-income families, which makes the Q--Q trade-off more
acute. In summary, Orbeta's study -- the first in the Philippines to
account for fertility endogeneity -- provides robust evidence that
increases in family size cause significant declines in child educational
attainment and validates the quantity--quality trade-off in this
context. Large family size thus emerges as one mechanism contributing to
poverty, by impeding children's human capital accumulation in the
Philippines.

@@

The Philippine experience is echoed in other ASEAN countries, where
researchers have probed the quantity--quality trade-off with diverse
outcomes and methods. In Vietnam, for example, rapid fertility decline
alongside rising education led to questions about a Q--Q mechanism.
\citet{anh1998family} documented a negative correlation between family
size and children's school enrollment in Vietnam, though their analysis
could not fully establish causality.

@@

Nonetheless, the prevailing evidence from the Philippines and its
regional neighbors strongly supports Becker's quantity--quality
conjecture. As families have fewer children, they appear to invest more
in each child's education---investments crucial for human capital
development and economic growth. Conversely, high-fertility households
risk under-investing per child, which reinforces cycles of poverty and
inequality. This literature thus provides an important empirical
foundation for policies in the Philippines and ASEAN---from family
planning programs to education subsidies---that aim to ease the
quantity--quality trade-off and help families achieve both manageable
size and better outcomes for the next generation.

@@

This study builds upon the foundational framework of the
quantity--quality (Q--Q) trade-off, as articulated in
\citet{becker1960economic} seminal work and later refined by
\citet{becker1973interaction}. Within this theoretical model, fertility
and child quality are jointly determined under a full-income constraint,
such that any increase in the number of children---absent a commensurate
expansion in resources---necessitates a reduction in the average
investment per child. While prior empirical applications of this
framework have largely centered on educational attainment, child quality
may also be understood along biological dimensions, especially along
nutritional status, which emerges earlier in the life course and may be
less subject to later remediation.

The current analysis focuses exclusively on early-life nutrition as a
proxy for child quality. Nutrition, unlike cognitive or educational
outcomes, is characterized by low substitutability and a narrow temporal
window during which investments are most productive. The trade-off is
thus expected to manifest strongly in settings where fertility rises
exogenously while household resources remain constrained. In such
environments, families facing an unexpected increase in sibship size may
be forced to reallocate scarce food, time, and health inputs across a
larger number of dependents, potentially resulting in lower nutritional
attainment for each child.

@@

In the presence of a fertility shock, holding \(I\) constant, the
average \(\kappa_j\) must fall unless offset by changes in parental
behavior or external transfers. If nutritional investments per child
decline, and if such inputs are critical for growth, observable deficits
in height-for-age (HAZ) or weight-for-height (WHZ) may emerge. The first
hypothesis emerging from this framework is that an exogenous increase in

fertility leads to a reduction in per-child nutritional investment,
particularly in low-income, urban households where slack in the family
budget is limited. A second, related hypothesis posits that this
reduction in inputs translates into measurable nutritional deficits
among young children, specifically in the form of lower anthropometric
outcomes such as height-for-age (HAZ) and weight-for-height (WHZ)
z-scores. These two hypotheses---concerning, respectively, the
behavioral reallocation of resources and its biological
consequence---jointly seek to identify a causal link between fertility
shocks and early-life deprivation.

@@

This study constructs a harmonized, child-level analytic dataset using
five rounds of the Philippine Demographic and Health Survey (DHS): 1993,
1998, 2003, 2008, and 2013. The DHS is a nationally representative,
cross-sectional survey series conducted by the Philippine Statistics
Authority in collaboration with ICF International. Each round follows a
stratified two-stage cluster design and collects standardized
information on child health, fertility, maternal characteristics, and
household infrastructure. For each wave, I merge the child-level (KR),
mother-level (IR), and household-level (HR) files using cluster,
household, and maternal identifiers, ensuring that each observation
corresponds to a live birth occurring within five years of the survey.
The analytic sample is restricted to children aged 0--59 months to
standardize developmental exposure windows and conform to the World
Health Organization (WHO) reference growth standards.

The primary outcomes are height-for-age (HAZ) and weight-for-height
(WHZ) z-scores, internationally recognized proxies for chronic and acute
malnutrition, respectively. These anthropometric indicators are cleaned
following DHS protocols: biologically implausible values are excluded,
and remaining z-scores are rescaled from hundredths to continuous units.
In addition to anthropometry, the child record includes a suite of
health-related variables. These include indicators for whether the child
has received basic vaccinations---such as BCG, DPT, polio, and
measles---as well as whether the child experienced episodes of diarrhea,
fever, or respiratory illness in the two weeks prior to the survey.
These supplementary health measures serve as intermediate outcomes in
secondary analyses to capture short-run morbidity and preventive health
investment.

Maternal-level information is drawn from the individual recode file.
Core covariates include maternal age, completed years of education, and
total number of children ever born (parity), each of which captures
socioeconomic status and fertility history. I additionally include
maternal body mass index (BMI) and anemia status, which serve as both
controls and potential mediators of child health. These indicators
reflect the mother's nutritional status and physiological capacity to
support healthy fetal and postnatal development. In robustness checks, I
examine whether declines in maternal nutritional status following the
fertility shock account for observed changes in child anthropometry.

Household-level characteristics are used to control for environmental
and infrastructural determinants of health. Specifically, I include
household size, urban residence, wealth quintile, access to electricity,
type of toilet facility, whether the toilet is shared, and source of
drinking water. These variables proxy for household-level resources and
exposure to disease environments. In extension analyses, I interact
treatment status with these environmental indicators to assess whether
the nutritional consequences of the fertility shock are magnified under
infrastructural deprivation.

The treatment is defined by exposure to Executive Order No.~003, which
was issued by the Manila City government in January 2000 and effectively
suspended the public distribution of modern contraceptives. This policy
change resulted in a localized increase in fertility that was not
mirrored in surrounding municipalities within the National Capital
Region (NCR). Children born in Manila City after the implementation of
the policy are classified as treated, while children in other NCR cities
serve as the control group. This spatially and temporally bounded
intervention provides quasi-experimental variation that facilitates
identification of causal effects.

The empirical approach adopts a difference-in-differences framework that
compares trends in child nutritional outcomes between treated and
control cities before and after the policy change. The identifying
assumption is that, absent the policy, nutritional trajectories in
Manila and non-Manila NCR cities would have evolved in parallel. This
assumption is supported by visual inspection and pre-trend placebo tests
using the 1993 and 1998 waves. The 1998 DHS serves as the primary
pre-policy baseline, while the 2003 survey captures short-run effects.
The 2008 and 2013 waves are used in robustness checks to assess medium-
and long-run persistence of nutritional deficits. All specifications
include city and birth-year fixed effects, and standard errors are
clustered at the city-year level to account for within-cluster
correlation.

The empirical strategy exploits the 2000 issuance of Executive Order
No.~003, which effectively banned the distribution of modern
contraceptives in public health centers across Manila City. This abrupt
and localized withdrawal of contraceptive access resulted in a fertility
increase concentrated within the city, but did not affect surrounding
municipalities within the National Capital Region (NCR). Children born
in Manila following the policy are considered exposed to a fertility
shock, while children in other NCR cities over the same period serve as
the control group. The analytical sample is restricted to children under
five years old at the time of each survey round to ensure accurate
measurement of early-life nutritional outcomes.

I estimate the following difference-in-differences specification:

\[ \text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict} \]

In this equation, \(\text{Nutrition}_{ict}\) refers to either
height-for-age (HAZ) or weight-for-height (WHZ) for child \(i\) in city
\(c\) born in year \(t\). The indicator \(\text{Post}_t\) equals 1 if
the child was born after the implementation of Executive Order No.~003
in 2000, and \(\text{Manila}c\) equals 1 if the child resides in Manila
City. The interaction term captures the effect of being born in Manila
after the contraceptive policy was enacted. The vector \(X{ict}\)
includes controls for child age in months, sex, birth order, maternal
age, education, parity, body mass index (BMI), anemia status, household
wealth quintile, household size, electricity access, water source,
toilet type, shared sanitation status, and urban residence. City fixed
effects \(\lambda_c\) control for time-invariant characteristics at the
locality level, while year fixed effects \(\tau_t\) absorb aggregate
time shocks. The coefficient of interest, \(\beta\), represents the
average treatment effect of the policy-induced fertility shock on child
nutrition. Standard errors are clustered at the city-year level to
account for potential autocorrelation within geographic units over time.

To support the internal validity of the design, I use Philippine Census
microdata from 1990, 1995, 2000, 2007, and 2010 to confirm the magnitude
and timing of the fertility shock. The 1995 and 2007 waves provide clean
pre- and post-policy benchmarks, while the 2000 census---although
temporally proximate to the policy---is retained for robustness checks
that test the sensitivity of estimates to the exact onset of fertility
changes. A placebo comparison between 1990 and 1995 shows no divergence
in fertility trends between Manila and control cities prior to the
policy, strengthening the credibility of the parallel trends assumption.

>>>>

\title{Womb for One More? Fertility Expansion and the Limits of Early
Health Investment: Empirical Evidence on the Quantity-Quality Trade-off
from a Contraceptive Ban in Urban Philippines}
\shortTitle{The Q-Q Trade-off in the Philippine Context}
% \author{Author1 and Author2\thanks{Surname1: affiliation1, address1, email1.
% Surname2: affiliation2, address2, email2. Acknowledgements}}

@@

The quantity-quality (Q--Q) theory, a central idea in modern family
economics, holds that parents face a trade-off between the number of
children and the ``quality'' of investment---such as education or
health---they can provide to each. Quality in this context refers to the
human capital of each child: attributes like education, health, and
nutrition that enhance a child's future productivity and well-being. The
genesis of this idea traces back to Gary Becker's seminal work around
1960, which for the first time treated children as economic goods
subject to parental choice and budget constraints
\citep{becker1960economic}. Becker argued that as families become
wealthier, they may not simply want more children, but rather
better-raised children, much as a household might prefer a
higher-quality car or house over a greater quantity of them. This
proposition led to a formal theory in which increases in income or
changes in economic conditions cause parents to substitute child quality
for quantity, consistent with historical patterns of lower fertility and
higher educational attainment during economic development
\citep{galor2000population}.

@@

Becker and Lewis (1973) formalize the mutually reinforcing nature of the
quantity-quality cost structure. An increase in \(n\) raises the total
cost required to sustain a given level of \(q\) for each child, while a
higher level of \(q\) raises the marginal cost associated with having an
additional child. For example, allocating more resources to education or
health per child increases the financial burden of expanding family
size. This interdependence links the two decisions directly. The
household cannot choose \(n\) and \(q\) in isolation; each choice alters
the marginal cost of the other.

@@

In many dynastic models, child quality appears indirectly, often through
the child's future human capital or income. One variant assumes parents
value the aggregate human‐capital stock of their children. This
specification, combined with altruism, produces a similar trade‐off:
concentrating resources in fewer children raises each child's human
capital, which raises the dynasty's long‐run welfare. These
intergenerational extensions link micro‐level fertility decisions to
macroeconomic outcomes. By the late 1980s, work by Becker, Barro, and
others had recast fertility as an endogenous choice that interacts with
capital accumulation, income distribution, and policy. This laid the
foundation for unified growth theories, which view the quantity-quality
mechanism as central to demographic transition and long-run development.

@@

The unified growth literature places the Q--Q model within a more
general account of economic and demographic change. In this framework,
higher income or stronger returns to child quality reduce fertility and
help shift economies from stagnation toward sustained growth. Several
mechanisms support this transition, such as a fall in child labor, a
drop in child mortality, and a shift in parental priorities. These
models explain not only the presence of a quantity-quality trade-off but
also its rising influence at a specific point in history. The evidence
supports these claims: countries that saw earlier increases in returns
to education experienced earlier fertility decline, while delays in
reforms, such as public education or health access, corresponded to
prolonged high fertility. As a result, the Q--Q mechanism forms a key
component of unified growth theory.

@@

Contemporary research has further refined the quantity-quality model by
relaxing some of its initial simplifying assumptions. Three important
extensions involve (1) capital market imperfections that constrain
parents' ability to invest in child quality, (2) intra-household
conflict and bargaining between mothers and fathers over fertility
choices, and (3) recognition that child quality is multi-dimensional,
which extends beyond schooling to include health, nutrition, and other
facets of human capital.

To begin, I examine how credit constraints can give rise to poverty
traps. The canonical quantity-quality (Q--Q) framework assumes that
parents can reallocate resources freely; borrowing against future
earnings to finance schooling or health investments whenever the
expected return is high. In practice, especially in low-income settings,
credit markets function imperfectly: poor households typically cannot
secure loans to cover children's education or medical costs even when
such investments would yield substantial future gains. This market
failure magnifies the Q--Q trade-off. Becker, Lewis, and Willis--already
noted by \citet{grawe2008quality} as emphasizing ``resource
limitations''--implicitly recognized this issue, but contemporary models
make it explicit by imposing a borrowing constraint. Parents must fund
childrearing from current income alone; they cannot collateralize a
child's future wages to pay today's school fees. Consequently, when
income is low, each additional birth directly reduces the attainable
quality per child, potentially trapping families in a
low-income--high-fertility equilibrium.

Formal models show that ``the quality-quantity trade-off arises from a
binding credit constraint that prevents parents from borrowing against
future child income.'' Empirical work supports this mechanism.
\citet{kremer2002income} and \citet{de2003inequality} document that
countries facing tighter liquidity constraints tend to display higher
fertility and lower educational attainment, consistent with
liquidity-constrained parents favoring quantity over quality.
Cross-country evidence likewise indicates that where financial frictions
are more severe, the negative correlation between fertility and
schooling is stronger. Theoretically, introducing a borrowing limit can
generate multiple steady states: one with low fertility and high
investment when incomes suffice to cover quality costs, and another with
high fertility and minimal investment when they do not. Policy
instruments such as education subsidies or conditional cash transfers
effectively relax these constraints, which nudge households toward the
low-fertility, high-investment equilibrium. In short, incorporating
credit market imperfections deepens the explanatory power of the Q-Q
model: economic growth alone may not reduce fertility if households
remain too cash-poor to afford schooling, whereas targeted
quality-enhancing transfers can catalyze both demographic and
human-capital transitions.

A further refinement of the quantity-quality (Q--Q) model considers the
question of \emph{who} within the household makes fertility and child
investment decisions. The original Beckerian framework adopts a unitary
model of the family, and assumes a single utility function and complete
agreement between spouses over optimal fertility \(n\) and child quality
\(q\). In practice, however, empirical evidence reveals significant
heterogeneity in preferences between household members along gender
lines. According to \citet{oppenheim1987impact, thomas1990intra}, for
instance, men often desire more children than women and may differ in
their willingness to invest in each child's education or health. These
discrepancies have motivated game-theoretic models of intra-household
bargaining, in which fertility and investment outcomes reflect the
relative influence of each parent's preferences.

@@

The original Q--Q models often used a single catch-all variable for
child quality, typically thought of as education or ``expenditure per
child.'' Recent work emphasizes that child quality is multi-faceted, and
that parents make trade-offs along several dimensions of
investment---cognitive development, health, nutrition, etc. This is
salient in developing countries, where basic health and nutrition are
pressing concerns alongside schooling.

@@

Beyond survival, early-life health indicators such as birth weight,
perceived size at birth, and vaccination status are important dimensions
of child quality, especially in low-income settings. When resources are
limited, parents may struggle to meet the basic needs of each child.
Empirical evidence aligns with the quantity--quality framework: children
from larger families often have worse health outcomes at birth and are
less likely to receive core immunizations. For example, recent
multi-country data show that the share of completely unimmunized
children---so-called ``zero-dose'' children---increases from 10.5\%
among those with no siblings to 17.2\% among those with four or more
\citep{costa2024child}. Studies from Southeast Asia also report that
higher birth order correlates with lower average birth weight and
smaller perceived size at birth. These patterns suggest that as
fertility increases, the health of each child may suffer due to more
limited parental investment. Studies from Southeast Asia
\citep{chen2021population} also report that higher birth order
correlates with lower average birth weight and smaller perceived size at
birth

Similar patterns emerge in sub-Saharan Africa
\citep{bishwakarma2019first} and South Asia, where first-born children
experience slower growth once younger siblings arrive---again pointing
to intrahousehold trade-offs. Theoretically, one can extend the Becker
model with a nutrition production function, where a child's health \(H\)
depends on food \(F\) and medical care \(M\), such that:

@@

Early empirical work in the Philippines provided suggestive support for
this theory. For example, an influential study by
\citet{horton1986child} used Philippine household data to jointly
examine fertility and child nutrition and treated nutritional status as
a measure of child quality. Horton found that better-educated mothers
and fathers tended to choose smaller families and achieved
better-nourished children--evidence of substitution away from
``quantity'' toward child ``quality.'' Notably, she observed strong
birth-order effects on nutrition (first-born children faring better than
later-born), which hint that parents may not evenly distribute quality
investments among all children. This early work indicated a
quantity-quality (Q--Q) trade-off in Philippine families, though it
largely documented correlations rather than definitive causation.

@@

The IV estimates confirmed a significant negative causal effect of
higher fertility on educational outcomes. \citet{orbeta2010number} found
that each additional child in the household reduced the proportion of
school-age children (6--24 years) attending school by roughly 19\% of
the baseline attendance rate. The trade-off was especially pronounced at
higher education levels: for example, the estimated drop in school
attendance was about 26\% at the secondary level and 57\% at the
tertiary level for each additional sibling. These are sizable effects,
implying that children from large families are substantially less likely
to remain in school, presumably due to tighter household budget
constraints or diluted parental attention. Moreover, the burden of the
trade-off appeared regressive: Orbeta's results showed much larger
schooling deficits from an extra child in poorer households than in
richer ones. For instance, in the poorest quintile, an additional
sibling reduced school attendance by an estimated 24\% (for ages 6--24),
compared to a 16\% reduction in the richest quintile. This regressive
pattern aligns with Becker's theory that resource constraints bind more
tightly for low-income families, which makes the Q--Q trade-off more
acute. In summary, Orbeta's study -- the first in the Philippines to
account for fertility endogeneity -- provides robust evidence that
increases in family size cause significant declines in child educational
attainment and validates the quantity-quality trade-off in this context.
Large family size thus emerges as one mechanism contributing to poverty,
by impeding children's human capital accumulation in the Philippines.

@@

The Philippine experience is echoed in other ASEAN countries, where
researchers have probed the quantity-quality trade-off with diverse
outcomes and methods. In Vietnam, for example, rapid fertility decline
alongside rising education led to questions about a Q--Q mechanism.
\citet{anh1998family} documented a negative correlation between family
size and children's school enrollment in Vietnam, though their analysis
could not fully establish causality.

@@

Nonetheless, the prevailing evidence from the Philippines and its
regional neighbors strongly supports Becker's quantity-quality
conjecture. As families have fewer children, they appear to invest more
in each child's education---investments crucial for human capital
development and economic growth. Conversely, high-fertility households
risk under-investing per child, which reinforces cycles of poverty and
inequality. This literature thus provides an important empirical
foundation for policies in the Philippines and ASEAN---from family
planning programs to education subsidies---that aim to ease the
quantity-quality trade-off and help families achieve both manageable
size and better outcomes for the next generation.

@@

This study builds upon the foundational framework of the
quantity-quality (Q--Q) trade-off, first articulated in
\citet{becker1960economic} and later refined in
\citet{becker1973interaction}. In this model, fertility and child
quality are jointly determined under a full-income constraint, such that
increases in the number of children---without a corresponding expansion
in household resources---necessitate a reduction in the average
investment per child. While empirical applications of this framework
have traditionally focused on educational outcomes, child quality can
also be expressed in terms of early-life health, including indicators
that emerge at birth or shortly after, such as birth weight, perceived
size at birth, and uptake of essential vaccinations.

This analysis focuses specifically on early-life health as a proxy for
child quality. These outcomes, unlike cognitive or schooling-related
investments, are biologically constrained and occur within a narrow
developmental window, during which parental inputs are least
substitutable and most consequential. The trade-off is expected to be
especially salient in settings where fertility increases exogenously
while family resources remain fixed. In such contexts, a sudden increase
in sibship size may compel households to spread food, healthcare, and
caregiving attention across more dependents, which would potentially
lead to diminished health investments and outcomes for each child.

@@

In the presence of a fertility shock, holding \(I\) constant, the
average \(\kappa_j\) must fall unless offset by changes in parental
behavior or external transfers. If these inputs decline and are critical
to early-life development, observable deficits may arise in birth weight
or perceived neonatal size, and children may be less likely to receive
complete immunization schedules.

The first hypothesis derived from this framework is that an exogenous
increase in fertility leads to a decline in per-child health investment,
particularly in low-income urban households with little financial slack.
A second, related hypothesis posits that this decline results in
measurable early-life health disadvantages---specifically, lower birth
weight, smaller perceived size at birth, and reduced uptake of key
childhood vaccinations. These hypotheses describe a causal pathway
through which fertility shocks may contribute to early deprivation and
potentially reinforce intergenerational cycles of poor health and
limited human capital.

@@

\subsection{Data}

The primary data source for this study is the Philippine Demographic and
Health Surveys (DHS), specifically the 1998, 2003, and 2008 rounds.
These nationally representative surveys provide detailed child- and
household-level information, including maternal characteristics,
fertility behavior, and early-childhood health outcomes. Importantly,
they are the only Philippine DHS waves with available GPS cluster data,
which I use to approximate geographic location and assign respondents to
treatment and control groups. Because city-level identifiers are not
released in the public-use files, I rely on coarse geographic assignment
based on cluster coordinates. Clusters located within or near the
administrative boundaries of the City of Manila are classified as
``treated,'' while clusters located in other parts of the National
Capital Region (NCR) serve as the control group. In robustness checks, I
restrict the control group to Quezon City or expand it to the rest of
NCR excluding both Manila and Quezon City.

The analytic sample includes children under five years of age living
with their mothers at the time of the survey. I focus on three
early-life health outcomes that reflect both biological endowments and
parental investment: (1) birth weight, reported in grams when available;
(2) the mother's subjective assessment of birth size; and (3)
vaccination uptake, defined as receipt of key early-childhood
immunizations, including BCG, DPT, polio, and measles. These outcomes
are consistently available across all three waves and serve as proxies
for child quality in the context of the quantity--quality trade-off.

The 1998 DHS is used as the primary pre-policy baseline wave because it
captures children born before the implementation of Manila's
contraceptive ban in 2000. The 2008 DHS serves as the post-policy round
to capture outcomes for children born well after the ban took effect. I
also include the 2003 DHS in the main specification, recognizing that
children in this round may have been born shortly before or after the
policy's introduction. I treat 2003 as a transition-period dataset and
conduct sensitivity checks to assess the robustness of findings to its
inclusion. I exclude the 1993 DHS from the analysis due to the absence
of GPS data, which makes it impossible to reliably assign treatment
status.

To support the identification strategy, I use DHS summary indicators on
contraceptive use disaggregated by region and city. These data confirm a
sharp and sustained decline in modern contraceptive use in Manila
following the implementation of Executive Order No.~003, relative to
surrounding cities, validating the policy as a plausibly exogenous
fertility shock. I also considered using data from the Philippine
National Nutrition Survey (NNS), which includes rich anthropometric and
dietary data. However, the NNS microdata are only available from 2003
onward, which limits their value for pre-policy analysis. As such, the
DHS remains the most suitable dataset for this study.

\subsection{Identification Strategy}

To estimate the causal effect of increased fertility on early-childhood
health outcomes, I exploit a policy shock generated by Executive Order
No.~003, issued in 2000 by the mayor of Manila. This order restricted
access to modern contraceptives in all city-run health facilities and
remained in effect for nearly a decade. The policy was implemented only
in the City of Manila and did not affect neighboring municipalities
within the National Capital Region (NCR), which continued to provide
modern family planning services. This spatially localized and time-bound
restriction created a natural experiment that allows for a
quasi-experimental research design. I define the treatment group as
children residing in GPS clusters located within or near the Manila
boundary, and the control group as children in other NCR municipalities.
Given the limitations of the DHS GPS data, which are displaced for
confidentiality and do not include exact city identifiers, I use a
coarser geographic classification to approximate exposure. Robustness
checks vary the control group to include Quezon City only, or the rest
of NCR excluding both Manila and Quezon City.

The core of my empirical strategy is a difference-in-differences (DiD)
design that compares changes in child outcomes in Manila before and
after the ban relative to changes in similar urban areas that were
unaffected. The key assumption is that, in the absence of the policy,
trends in early-childhood health outcomes would have followed similar
trajectories across treatment and control areas. This approach accounts
for both fixed differences between cities and nationwide time trends.

To address potential violations of the parallel trends assumption---such
as Manila-specific shocks that may have affected all children regardless
of family size---I implement a triple-differences strategy that
leverages variation in exposure by maternal age. Because fertility
declines sharply with age, older women were biologically less likely to
respond to the contraceptive ban. For example, mothers aged 45--49 at
the time of the policy were near the end of their reproductive years and
thus unlikely to experience a fertility increase. By comparing changes
in child outcomes across maternal age cohorts within Manila and control
cities, I can account for time-varying, location-specific confounders
that affect all families, while isolating the policy's effect on family
size and its consequences.

I further restrict my main analysis to children born before the policy
was implemented. Although these children were not conceived due to the
ban, they may still have been indirectly affected by the fertility
increase through resource dilution within households. This strategy
avoids selection bias related to unobservable differences in the
endowments of ``marginal children'' born because of the policy. As
argued by Rosenzweig and Zhang (2009), when parents reinforce rather
than compensate for birth endowments, estimates based on older siblings
provide a lower bound of the true effect of fertility on child quality.
Empirical evidence from the Philippines supports this pattern of
reinforcement, as shown by Baguet and Dumas (2015). Taken together,
these strategies allow me to estimate the causal impact of fertility
shocks on early-life health while addressing key concerns about internal
validity and sample composition.

\subsection{Sample Selection}

The analytic sample is drawn from the Philippine Demographic and Health
Surveys (DHS), specifically the Kid Recode (KR) files from 1998, 2003,
and 2008. I restrict the sample to children under age five (0--59
months) with non-missing information on key outcomes: birth weight (in
grams, when available), perceived size at birth, and vaccination status.
Because the Manila contraceptive ban was enacted in 2000 and likely
implemented unevenly at first, I define children born in 2001 and later
as post-policy and those born before 1997 as pre-policy. Children born
between 1997--2000 are excluded in the main specification but retained
in robustness checks.

I include all live births regardless of birth order or plurality to
preserve variation in sibship size. Twins and higher-order multiple
births are flagged for sensitivity checks but not excluded. Each child
is linked to their biological mother using maternal identifiers,
allowing construction of sibship size, birth order, and maternal age.
Exposure to the policy is defined geographically: children are
considered treated if they were born in the City of Manila, and
untreated if born in other municipalities within the National Capital
Region (NCR). Because GPS data are coarse in some years, geographic
exposure is defined at the city level rather than using precise
coordinates. Households outside NCR are excluded from the main analysis
to ensure urban comparability and avoid confounding by rural fertility
and health dynamics.

\subsection{Variables}

The primary outcome variables reflect early-life health and investment.
These include: (1) birth weight in grams, as recorded by the mother when
available; (2) the mother's subjective assessment of birth size
(categorized as very small, small, average, large, or very large); and
(3) indicators for whether the child received core early-life
vaccinations, including BCG, DPT1--3, polio doses, and measles. A binary
variable indicating whether the child received all recommended vaccines
is also constructed as a summary measure of postnatal health investment.
